MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-22 docs/paper/out/paper.tex
Added v- prefix for the versions; some lost dots

The exponential growth of metagenomics is adding a significant plus of
complexity to the big data problem in genomics. In this new scenario
impacted by the wide scale and scope of the projects and by the
explosion of sequence data to be analyzed is especially opportune the
use of new possibilities that cloud computing approaches, new functional
and dependently typed programming languages and new database paradigms
as graph databases offer. To tackle the challenges of big data analysis
in this work we have used these new means to design and develop a new
open source methodology for analyzing metagenomics data, MG7. It
exploits the new possibilities that cloud computing offers to get a
system robust, programmatically configurable, modular, distributed,
flexible, scalable and traceable in which the biological databases of
reference sequences can be easily updated and / or frequently
substituted by new ones or by databases specifically designed for
focused projects. MG7 uses parallelization and distributed analysis
based on AWS, with on-demand infrastructure as the basic paradigm and
allow the definition of complex workflows using a composable system for
scaling/parallelizing stateless computations designed for Amazon Web
Services (AWS) that counts with a static reproducible specification of
dependencies and behavior of the different components. The modelling of
the taxonomy tree is done using the new paradigm of graph databases of
Bio4j that facilitates the taxonomic assignment tasks and the
calculation of the taxa abundance values considering the hierarchic
structure of taxonomy tree. MG7 includes the new 16S database 16S-DB7
built with a flexible and sustainable system of updating and
project-driven personalization.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Sequence data:} the reads are larger, the sequencing depth and
  the number of samples of each project are considerably bigger. The
  first metagenomics studies were very local projects, while nowadays
  the most fruitful studies are done at a global level (international,
  continental, national). This kind of global studies has yielded the
  discovery of clinical biomarkers for diseases of the importance of
  cancer, obesity or inflammatory bowel diseases and has allowed
  exploring the biodiversity of varied earth environments.
\item
  \textbf{The genomics explosion:} its effect being felt in this case in
  the reference sequences. The immense amount of sequences available in
  public repositories demands new strategies for curation, update and
  storage of metagenomics reference databases: current models will
  (already) have problems to face the future avalanche of metagenomic
  sequence data.
\item
  \textbf{Cloud computing:} the appearance of new models for massive
  computation and storage such as the so-called cloud, or the widespread
  adoption of programming methodologies like functional programming, or,
  more speculatively, dependently typed programming. The immense new
  possibilities that these advances offer must have a direct impact in
  metagenomics data analysis.
\item
  \textbf{Open science:} the new social manner to do science,
  particularly so in genomics, brings its own set of requirements.
  Metagenomics evolves in a social and global scenario following a
  science democratization trend in which many small research groups from
  distant countries share a common big metagenomics project; this global
  cooperation demands systems allowing for reproducible data analysis,
  data interoperability, and tools and practices for asynchronous
  collaboration between different groups.
\end{enumerate}

@@

\href{https://github.com/ohnosequences/datasets}{Datasets} is a Scala
library developed by the first and last authors with the goal of being a
Scala-embedded mini-language for datasets and their locations.
\textbf{Data} is represented as type-indexed fields: Keys are modeled as
singleton types, and values correspond to what could be called a
denotation of the key: a value of type \texttt{Location} tagged with the
key type. Then a \textbf{Dataset} is essentially a collection of data,
which are guaranteed statically to be different through type-level
predicates, making use of the value ↔ type correspondence which can be
established through singleton types and implicits. A dataset location is
then just a list of locations formed by locations of each data member of
that dataset. All this is based on what could be described as an
embedding in Scala of an extensible record system with concatenation on
disjoint labels, in the spirit of
\citep[\citet{harper1991record}]{harper1990extensible}. For that
\emph{Datasets} uses
\href{https://github.com/ohnosequences/cosas/}{ohnosequences/cosas}.

@@

We developed our own Scala-based type-safe eDSLs (\textbf{e}mbedded
\textbf{D}omain \textbf{S}pecific \textbf{L}anguage) for
\href{https://github.com/ohnosequences/flash}{FLASH} and
\href{https://github.com/ohnosequences/blast}{BLAST} expressions and
their execution.

In the case of BLAST we use a model for expressions where we can
guarantee for each BLAST command expression at compile time

@@

Generic type-safe parsers returning a heterogeneous record of BLAST
output fields are also available, together with output data defined
using \emph{Datasets} which have a reference to the exact BLAST command
options which yielded that output. This let us provide generic parsers
for BLAST output which are guaranteed to be correct.

@@

Certainly, if metagenomics results are easily integrated with the
theoretical and experimental knowledge of each specific area, the impact
of metagenomics will be higher that if metagenomics progresses as a
disconnected research branch. Considering that metagenomics data
interoperability, which is especially critical in clinical environments,
requires a stable taxonomy to be used as reference, we decided to rely
on the most widely used taxonomy: the NCBI taxonomy. In addition, the
biggest global sequence database GenBank follows this taxonomy to
register the origin of all their submitted sequences. Our 16S database
building strategy allows the substitution of the 16S database by any
other subset of \textbf{nt}, even by the complete \textbf{nt} database
if it would be needed, for example, for analyzing shotgun metagenomics
data. This possibility of changing the reference database provides
flexibility to the system enabling it for easy updating and
project-driven personalization.

@@

MG7 proposes to define and work with a particular data analysis task as
a software project, using Scala. The idea is that \emph{everything}:
data description, their location, configuration parameters, the
infrastructure used, \ldots{} should be expressed as Scala code, and
treated in the same way as any (well-managed) software project. This
includes, among other things, using version control systems
(\texttt{git} in our case), writing tests, making stable releases
following \href{http://semver.org/}{semantic versioning} or publishing
artifacts to a repository.

@@

These services are used through a Scala wrapper of the official
\href{https://aws.amazon.com/sdk-for-java/}{AWS Java SDK
\texttt{1.9.25}}:
\href{https://github.com/ohnosequences/aws-scala-tools/releases/tag/v0.13.2}{ohnosequences/aws-scala-tools
\texttt{0.13.2}}.

@@

MG7 itself and all the libraries used are written in Scala
\texttt{2.11}.

@@

MG7 uses
\href{https://github.com/statika/statika/releases/tag/v0.2.0}{ohnosequences/statika
\texttt{2.0.0}} for specifying the configuration and behavior of EC2
instances.

@@

MG7 uses
\href{https://github.com/ohnosequences/datasets/releases/tag/v0.2.0}{ohnosequences/datasets
\texttt{0.2.0}} for specifying input and output data, their type and
their location.

@@

MG7 uses
\href{https://github.com/ohnosequences/loquat/releases/tag/v2.0.0}{ohnosequences/loquat
\texttt{2.0.0}} for the specification of data processing tasks and their
execution using AWS resources.

@@

MG7 uses
\href{https://github.com/ohnosequences/blast/releases/tag/v0.2.0}{ohnosequences/blast
\texttt{0.2.0}}. The BLAST version used is \texttt{2.2.31+}

@@

MG7 uses
\href{https://github.com/ohnosequences/flash/releases/tag/v0.1.0}{ohnosequences/flash
\texttt{0.1.0}}. The FLASH version used is \texttt{1.2.11}

@@

MG7 uses
\href{https://github.com/bio4j/bio4j/releases/tag/v0.12.0-RC3}{bio4j/bio4j
0.12.0-RC3} and
\href{https://github.com/bio4j/bio4j-titan/releases/tag/v0.4.0-RC2}{bio4j/bio4j-titan
0.4.0-RC2} as an API for the NCBI taxonomy.

>>>>

The exponential growth of metagenomics is adding a significant plus of
complexity to the big data problem in genomics. In this new scenario
impacted by the wide scale and scope of the projects and by the
explosion of sequence data to be analyzed is especially opportune the
use of new possibilities that cloud computing approaches, new functional
and dependently typed programming languages and new database paradigms
as graph databases offer. To tackle the challenges of big data analysis
in this work we have used these new means to design and develop a new
open source methodology for analyzing metagenomics data, MG7. It
exploits the new possibilities that cloud computing offers to get a
system robust, programmatically configurable, modular, distributed,
flexible, scalable and traceable in which the biological databases of
reference sequences can be easily updated and/or frequently substituted
by new ones or by databases specifically designed for focused projects.
MG7 uses parallelization and distributed analysis based on AWS (Amazon
Web Services), with on-demand infrastructure as the basic paradigm and
allow the definition of complex workflows using a composable system for
scaling/parallelizing stateless computations designed for AWS that
counts with a static reproducible specification of dependencies and
behavior of the different components. The modelling of the taxonomy tree
is done using the new paradigm of graph databases of Bio4j that
facilitates the taxonomic assignment tasks and the calculation of the
taxa abundance values considering the hierarchic structure of taxonomy
tree. MG7 includes the new 16S-DB7 database built with a flexible and
sustainable system of updating and project-driven personalization.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Sequence data:} the reads are larger, the sequencing depth and
  the number of samples of each project are considerably bigger. The
  first metagenomics studies were very local projects, while nowadays
  the most fruitful studies are done at a global level (international,
  continental, national). This kind of global studies has yielded the
  discovery of clinical biomarkers for diseases of the importance of
  cancer, obesity or inflammatory bowel diseases and has allowed
  exploring the biodiversity of varied earth environments.
\item
  \textbf{The genomics explosion:} its effect being felt in this case in
  the reference sequences. The immense amount of sequences available in
  public repositories demands new strategies for curation, update and
  storage of metagenomics reference databases: current models will
  (already) have problems to face the future avalanche of metagenomic
  sequence data.
\item
  \textbf{Cloud computing:} the appearance of new models for massive
  computation and storage such as the so-called cloud, or the widespread
  adoption of programming methodologies like functional programming, or,
  more speculatively, dependently typed programming. The new
  possibilities that these advances offer must have a direct impact in
  metagenomics data analysis.
\item
  \textbf{Open science:} the new social manner to do science,
  particularly so in genomics, brings its own set of requirements.
  Metagenomics evolves in a social and global scenario following a
  science democratization trend in which many small research groups from
  distant countries share a common big metagenomics project; this global
  cooperation demands systems allowing for reproducible data analysis,
  data interoperability, and tools and practices for asynchronous
  collaboration between different groups.
\end{enumerate}

@@

\href{https://github.com/ohnosequences/datasets}{Datasets} is a Scala
library developed by the first and last authors with the goal of being a
Scala-embedded mini-language for datasets and their locations.
\textbf{Data} is represented as type-indexed fields: keys are modeled as
singleton types, and values correspond to what could be called a
denotation of the key: a value of type \texttt{Location} tagged with the
key type. Then a \textbf{Dataset} is essentially a collection of data,
which are guaranteed statically to be different through type-level
predicates, making use of the value--type correspondence which can be
established through singleton types and implicits. A dataset location is
then just a list of locations formed by locations of each dataset key.
All this is based on what could be described as an embedding in Scala of
an extensible record system with concatenation on disjoint labels, in
the spirit of \citep[\citet{harper1991record}]{harper1990extensible}.
For that \emph{Datasets} uses
\href{https://github.com/ohnosequences/cosas/}{ohnosequences/cosas}
library.

@@

We developed our own Scala-based type-safe eDSLs (embedded Domain
Specific Languages) for
\href{https://github.com/ohnosequences/flash}{FLASH} and
\href{https://github.com/ohnosequences/blast}{BLAST} expressions and
their execution.

In the case of BLAST we use a model where we can guarantee for each
BLAST command expression at compile time that

@@

Generic type-safe parsers returning a heterogeneous record of BLAST
output fields are also available, together with output data defined
using \emph{Datasets} which have a reference to the exact BLAST command
options which yielded that output. This lets us provide generic parsers
for BLAST output which are guaranteed to be correct.

@@

Certainly, if metagenomics results are easily integrated with the
theoretical and experimental knowledge of each specific area, the impact
of metagenomics will be higher than if it progresses as a disconnected
research branch. Considering that metagenomics data interoperability,
which is especially critical in clinical environments, requires a stable
taxonomy to be used as reference, we decided to rely on the most widely
used taxonomy: the NCBI taxonomy. In addition, the biggest global
sequence database GenBank follows this taxonomy to register the origin
of all their submitted sequences. Our 16S database building strategy
allows the substitution of the 16S database by any other subset of
\textbf{nt}, even by the complete \textbf{nt} database if it would be
needed, for example, for analyzing shotgun metagenomics data. This
possibility of changing the reference database provides flexibility to
the system enabling it for easy updating and project-driven
personalization.

@@

MG7 proposes to define and work with a particular data analysis task as
a software project, using Scala. The idea is that \emph{everything}:
data description, their location, configuration parameters and the
infrastructure used should be expressed as Scala code, and treated in
the same way as any (well-managed) software project. This includes,
among other things, using version control systems (\texttt{git} in our
case), writing tests, making stable releases following
\href{http://semver.org/}{semantic versioning} or publishing artifacts
to a repository.

@@

These services are used through a Scala wrapper of the official
\href{https://aws.amazon.com/sdk-for-java/}{AWS Java SDK v1.9.25}:
\href{https://github.com/ohnosequences/aws-scala-tools/releases/tag/v0.13.2}{ohnosequences/aws-scala-tools
v0.13.2}.

@@

MG7 itself and all the libraries used are written in Scala v2.11.

@@

MG7 uses
\href{https://github.com/statika/statika/releases/tag/v0.2.0}{ohnosequences/statika
v2.0.0} for specifying the configuration and behavior of EC2 instances.

@@

MG7 uses
\href{https://github.com/ohnosequences/datasets/releases/tag/v0.2.0}{ohnosequences/datasets
v0.2.0} for specifying input and output data, their type and their
location.

@@

MG7 uses
\href{https://github.com/ohnosequences/loquat/releases/tag/v2.0.0}{ohnosequences/loquat
v2.0.0} for the specification of data processing tasks and their
execution using AWS resources.

@@

MG7 uses
\href{https://github.com/ohnosequences/blast/releases/tag/v0.2.0}{ohnosequences/blast
v0.2.0}. The BLAST version used is v2.2.31+.

@@

MG7 uses
\href{https://github.com/ohnosequences/flash/releases/tag/v0.1.0}{ohnosequences/flash
v0.1.0}. The FLASH version used is v1.2.11.

@@

MG7 uses
\href{https://github.com/bio4j/bio4j/releases/tag/v0.12.0-RC3}{bio4j/bio4j
v0.12.0-RC3} and
\href{https://github.com/bio4j/bio4j-titan/releases/tag/v0.4.0-RC2}{bio4j/bio4j-titan
v0.4.0-RC2} as an API for the NCBI taxonomy.
